### 2025-04-27 paper/paper.tex
Update paper.tex

Our neural network is implemented through the python libraries traffic control interface (traci) and pytorch. Traci allows us to modify our simulation and get state information within a current timestep, while pytorch allows us to easily build and implement neural networks for our project.

The first step in designing the proposed neural network is to establish what an agent is, and how they will communicate with the model. Based on existing codebases and research, we decided to implement a decentralized multi-agent reinforcement learning model. This specific implementation lacks a central coordinator, and each agent acts on its own in coordination with other agents. The agents were designed to represent an intersection with a traffic signal controller, and could accept the state of the lanes leading to the intersection, the state of the signals, and the states of vehicles at each intersection such as stopped time; it also accepted a list of possible actions the traffic signal controller could perform, which when provided to a deep-q learning model can guess the optimal move.

After designing an agent we decided to model the simulation environment in python to actually obtain states and actions for each agent. To do this, we used traci to obtain the states of vehicles on the lanes of traffic heading towards a specific traffic light, and assigned them to the traffic light. We then grabbed the possible state of actions, again assigning it to said traffic signal. This is necessary for computing optimal actions later. We gather these every 5 seconds in the simulation, which greatly increases performance, and prevents rapidly switching signals during learning.

Our initial plan was to predict the optimal states of every intersection with a traffic light in Starkville, MS, however, upon implementation we realized our available hardware and time frame would not allow us to do this. Instead, we opted to only use what we decided were points of interest, or intersections where traffic is particularly backed up. Most of these points were decided by considering the importance of the road, the time spent at intersections, and personal experience. This way we don't waste computational resources at intersections that see little traffic, or do fine directing traffic on their own. Doing so significantly decreases the computational requirements to run the model, and also allows us to focus only on important points.

Our model does not have preset correct moves, which makes designing a neural network a very difficult task. Therefore, its important to choose a (non supervised), q-score, action, learn.

implementing benchmarks

>>>>

Our decentralized multi‐agent reinforcement learning (MARL) architecture is built around three key components: the Q‐network itself, a DQN‐based agent wrapper (complete with replay buffer and learning logic), and a SUMO–traci environment interface that supplies state and reward information every control interval. Traci allows us to easily modify our simulation and get state information from within the current timestep, while pytorch allows us to efficiently build and implement our neural networks. This implementation lacks a central coordinator. Each agent acts on its own in coordination with other agents. The agents were designed to represent an intersection with a traffic signal controller, and could accept the state of the lanes leading to the intersection, the state of the signals, and the states of vehicles at each intersection such as stopped time. We define the neural network in models.py as a simple multilayer perceptron. Given an input dimension equal to the total number of lanes feeding into a signalized intersection and an output dimension equal to the number of possible signal phases, we stack two hidden layers of configurable width (48 units each in our experiments) with ReLU activations, and terminate with a linear layer producing a Q‐value for each phase.

Concretely, the helper function
\begin{verbatim}
build_mlp(input_dim, output_dim, hidden_layers):
    layers   = []
    prev     = input_dim
    for h in hidden_layers:
        layers.append(nn.Linear(prev, h))
        layers.append(nn.ReLU())
        prev = h
    layers.append(nn.Linear(prev, output_dim))
    return nn.Sequential(*layers)
\end{verbatim}
constructs the sequential network, allowing us to easily swap hidden‐layer sizes or depths via configuration.

Each intersection is managed by an instance of {\tt DQNAgent} (in policies.py), which maintains both an \emph{online} network and a \emph{target} network. At initialization, we load the target’s weights from the online network, and set up an Adam optimizer (learning rate 0.001), a discount factor $\gamma=0.99$, and an $\varepsilon$‐greedy exploration schedule that decays $\varepsilon$ from 1.0 to 0.05 by a factor of 0.995 per learning step. Experiences are stored in a fixed‐capacity deque buffer (size 5,000), and we sample mini‐batches of 32 transitions once the buffer is sufficiently large. In each learn() call we compute the predicted Q‐values for state–action pairs, form TD‐targets via \(Q_{\mathrm{target}} = r + \gamma (1 - d)\max_{a'} Q_{\mathrm{target}}(s', a')\) and minimize the mean‐squared error between predictions and targets. After each gradient update we decay $\varepsilon$, and every 100 updates we synchronize the target network with the online network. To scale this to multiple intersections, we wrap several {\tt DQNAgent} objects in {\tt MultiDQNAgent}, forwarding action selection, memory storage, and learning calls independently to each agent.

On the environment side (multi\_env.py, subclassing BaseEnv), we employ the python library traci (traffic control interface) to simulate traffic in Starkville, MS. Our initial plan was to predict the optimal states of every intersection with a traffic light in Starkville, MS, however, upon implementation we realized our available hardware and the time frame would not allow us to do this. Instead, we opted to only use what we decided were points of interest, or intersections where traffic is particularly heavy and prone to congestion. Most of these points were decided by considering the importance of the road, the typical time spent at intersections, and personal experience. This method ensures that we don't expend computational resources at intersections that either see little traffic or do fine managing traffic on their own. Doing so significantly decreases the computational requirements to run the model, and allows us to focus only on critical points. At startup we invoke traci.start with a step length of 0.1s and a control interval of 5s, disabling verbose logs to save overhead. We automatically collect, for each traffic‐light signal (TLS), the set of incoming lanes by inspecting traci.trafficlight.getControlledLinks. Every control step, get\_state() returns a list of queue lengths, i.e.\ the number of halting vehicles per lane, by calling traci.lane.getLastStepHaltingNumber on each lane. We then compute a scalar reward for each TLS as the negative sum of its queue lengths, which incentivizes the agents to minimize congestion. When step(actions) is called, we apply each agent’s chosen phase via traci.trafficlight.setPhase, advance the simulation by the control interval, and then return the new state, reward list, and a done flag (triggered when no vehicles remain).

The overall training loop lives in main.py, which begins by parsing our configuration file. This configuration specifies 500 training episodes, an agent name ``multi\_dqn,'' and all key hyperparameters (learning rate, $\gamma$, $\varepsilon$ schedule, hidden‐layer sizes, buffer and batch sizes, and target‐update frequency). For each episode, we reset the environment, then for up to 2,400 steps (equivalent to 0.75 hours of simulated time) we query the multi‐agent for actions, step the environment, store transitions, and trigger a learning update immediately. Episode returns (the sum of all TLS rewards) are logged each episode, and upon completion of training we save each agent’s weights to disk. A similar loop runs in evaluation mode (with greedy action selection and no learning) to report average performance over ten episodes.
